Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-08-30 Manuscript/manuscript.Rmd
intro edits

A growing number of studies have proposed computational tools used to identify the complete set of key molecular drivers of the formation of TAD boundaries. The classical approach is to build classification models that leverage the relationship between ChIP-seq genomic annotations and TAD boundary regions. A key factor in these approaches becomes resolution. Resolution determines the bin sizes of the linear genome used to enumerate contacts caused by chromatin folding. Due to high sequencing cost, most available Hi-C datasets have relatively low resolution, in the kilobase range. This is in stark contrast to ChIP-seq experiments which annotate regions of the genome at hundreds of base pairs in resolution [??? REF, Bonev 2017?]. This resolution gap between 1D and 3D genomic maps, as well as a paucity of chromatin features below the level of TADs, has significantly limited our understanding of epigenomic regulation its potential link to chromatin architecture. However, simply increasing sequencing depth to obtain higher resolution could have unintended consequences since this could introduce sparsity in the number of contacts which TAD callers are often ill-equiped to handle [@cresswell2019spectraltad, @sexton2012three]. Increased resolution could also further exacerbate the class imbalance created from the proportionally much smaller number of TAD boundary regions compared to non-TAD boundary regions throughout the linear genome. Classifiers implemented on heavily imbalanced datasets are known to exhibit poor performances as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. It is unclear how simple data level solutions to class imbalance, such random under-sampling and random over-sampling affect TAD boundary prediction. Additionally, how relationships between genomic anntotations and bins of varying resolutions are quantified remains an open question. Incorporating a distance measure has largely been ignored, but could offer a much more spatially descriptive feature space thereby improving predictive performance.

Recently, an L1-regularized multiple logistic regression model, called *HiCfeat*, 
was developed to predict TAD boundaries using transcription factor binding sites (TFBS) 
on GM12878 [@mourad2016computational]. High resolution Hi-C data at 1 kb was used and percentage of overlap between regions defining TFBS and 1 kb bins was used for feature engineering. However, class imbalance was not adressed and, instead area under the receiver operating curve (AUROC) was used to evaluate model performance. This is problematic since it has been shown that AUC is insensitive to class imbalance, creating artificially inflated values inlfuenced by the majority class [@prati2011survey; @wei2013role]. Two additional studies were proposed in 2015 and 2017 respectively, one using histone modifications in a Bayesian Additive Regression Trees (BART) model [@huang2015predicting] and the other using combined sets of TFBS, DNase hypersensitive sites, and histone modifications in an L2-based regularized linear model [@hong2017computational]. Both used re-sampling techniques prior to modeling, with both implementing random under-sampling to balance TAD and non-TAD boundary regions. However, given the large disparity between the number of genomic regions labeled as containing a TAD boundary or not, performing random under-sampling alone is likely to be unstable and introduce bias. Also, it is unclear how other data level solutions compare to simple random under-sampling such as random over-sampling or a weighted combination of both random under- and over-sampling together. Furthermore, there was not a consensus in either resolution or feature engineering. The BART model was implemented on 5 kb Hi-C data using sequence read count between ChIP-seq data and 5 kb bins, while the L2-based regularized linear model was implemented on filtered TADs greater than 200 kb in length using the elemental count that appeared in each 150 kb bin.

## OUR APPROACH

In this study, we have developed a computational framework aimed at determining the optimal set of data level characteristics necessary for optimal prediction of TAD boundaries and identification of influential genomic annotations. These include the best combination of predictor type, resampling technique, and data resolution. We utilize random forest classification models built on Hi-C data from two cell lines, including GM12878 and K562, obtained at four different resolutions, including 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For each cell line and resolution, we compared three predictor types, including counts, percents, and distances. Additionally, we compared three re-sampling techniques, including random under-sampling, random over-sampling, and a weighted combination of both referred to as Synthetic Minority Over-sampling Technique (SMOTE). Model performances were evaluated and compared using Matthews Correlation Coefficient (MCC). We performed 10-fold cross-validation for parameter tuning and to guard against over-fitting. We found that the optimally built cell line specific model is one implemented using distance type predictors, with SMOTE resampling, at the 10-kb data resolution. We also determined that simply including transcription factor binding sites was necessary for optimal performance. The inclusion of chromatin states and histone modifications was found to induce noise into the model. Specifically, we found that the transcription factors CTCF, ZNF143, RAD21, and SMC3 exhibited high predictive importances, confirming there involvement in TAD boundary formation. We then aimed to evaluate the cell line specificity of TAD boundaries by validating our optimally built models across cell lines. We found that models built and validated on the same cell line outperformed models that were validated across cell lines, confirming the cell line specific nature and plasticity of TAD boundaries and functional genomic annotations respectively. <!--[add conclusion from results of predicting finer bins]-->

>>>>

A growing number of studies have proposed computational tools used to identify the complete set of key molecular drivers of the formation of TAD boundaries. The classical approach is to build classification models that leverage the relationship between ChIP-seq genomic annotations and TAD boundary regions [@mourad2016computational; @huang2015predicting; @hong2017computational]. A key factor in these approaches becomes resolution. Resolution determines the bin sizes of the linear genome used to enumerate contacts caused by chromatin folding. Due to high sequencing cost, most available Hi-C datasets have relatively low resolution, in the kilobase range. This is in stark contrast to ChIP-seq experiments which annotate regions of the genome at hundreds of base pairs in resolution [??? REF, Bonev 2017?]. This resolution gap between 1D and 3D genomic maps, as well as a paucity of chromatin features below the level of TADs, has significantly limited our understanding of epigenomic regulation its potential link to chromatin architecture. However, simply increasing sequencing depth to obtain higher resolution could have unintended consequences since this could introduce sparsity in the number of contacts which TAD callers are often ill-equiped to handle [@cresswell2019spectraltad, @sexton2012three]. Increased resolution could also further exacerbate the class imbalance created from the proportionally much smaller number of TAD boundary regions compared to non-TAD boundary regions throughout the linear genome. Classifiers implemented on heavily imbalanced datasets are known to exhibit poor performances as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. It is unclear how simple data level solutions to class imbalance, such random under-sampling and random over-sampling affect TAD boundary prediction. Additionally, identifying the most informative feature engineering steps used to quantify the relationships between genomic anntotations and bins of varying resolutions remains an open question. Incorporating a distance measure has largely been ignored, but could offer a much more spatially descriptive feature space thereby improving predictive performance.

<!--
Recently, an L1-regularized multiple logistic regression model, called *HiCfeat*, 
was developed to predict TAD boundaries using transcription factor binding sites (TFBS) 
on GM12878 [@mourad2016computational]. High resolution Hi-C data at 1 kb was used and percentage of overlap between regions defining TFBS and 1 kb bins was used for feature engineering. However, class imbalance was not adressed and, instead area under the receiver operating curve (AUROC) was used to evaluate model performance. This is problematic since it has been shown that AUC is insensitive to class imbalance, creating artificially inflated values inlfuenced by the majority class [@prati2011survey; @wei2013role]. Two additional studies were proposed in 2015 and 2017 respectively, one using histone modifications in a Bayesian Additive Regression Trees (BART) model [@huang2015predicting] and the other using combined sets of TFBS, DNase hypersensitive sites, and histone modifications in an L2-based regularized linear model [@hong2017computational]. Both used re-sampling techniques prior to modeling, with both implementing random under-sampling to balance TAD and non-TAD boundary regions. However, given the large disparity between the number of genomic regions labeled as containing a TAD boundary or not, performing random under-sampling alone is likely to be unstable and introduce bias. Also, it is unclear how other data level solutions compare to simple random under-sampling such as random over-sampling or a weighted combination of both random under- and over-sampling together. Furthermore, there was not a consensus in either resolution or feature engineering. The BART model was implemented on 5 kb Hi-C data using sequence read count between ChIP-seq data and 5 kb bins, while the L2-based regularized linear model was implemented on filtered TADs greater than 200 kb in length using the elemental count that appeared in each 150 kb bin. 
-->

To date, there has yet to be established a unified approach toward TAD boundary prediction. Therefore, in this study we first developed a computational framework aimed at determining the set of data level characteristics necessary for optimal TAD boundary prediction. We assessed the Matthews Correlation Coefficient (MCC) of 10-fold cross-validated random forest classifiers with different combinations of model inputs consisting of vary data resolutions, re-sampling techniques, and feature engineering steps. We showed that 10 kb resolution data, combined with both random over- and under-sampling, and a feature space composed of distances between ChIP-seq annotations and genomic bins yielded the best model performances. We showed that our optimally built model was able to identify known architectural proteins CTCF, RAD21, SMC3, and ZNF143 as exhibiting high predictive importance, confirming their biological functionality near TAD boundaries. We then aimed to use our optimally built model to more precisely predict the location of TAD boundaries at base pair resolution. We showed that, compared to TAD boundaries identified by a TAD caller, these top ranked transcription factors were much more enriched near the predicted TAD boundaries identified by our model indicating much more precise locations throughout the linear genome. This improved precision in TAD boundary location can provide insight into the role that epigenomics has on the spatial architecture of the human genome
